Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-03-06 paper/paper_orig_submission.tex
PLoS Pathogens reviewer response

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.
Here we develop a new method to assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by determining both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are often initiated by virions with defects that include amino-acid mutations, internal deletions, and failure to express key genes.
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
Most cells are infected by virions that have a mutation or some other defect.
This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and some viral mutations make a cell more likely to activate an innate-immune response.
However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and produce interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor looms large: viral genetic diversity.
Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if there are mutations in the virion infecting any given cell.

Here we develop a new approach to measure both the transcriptome and full sequences of all viral genes in single influenza-infected cells.
To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Two-thirds of cells are infected by virions with a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
We identify several types of viral defects that increase IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection*{A system to identify and enrich rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

A challenge in studying IFN induction by influenza virus is its rareness at the level of single cells.
Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction occurs at a frequency of $<$1\% in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).

To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
For the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

\caption{
{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
%From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets along with low-quality droplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts as is commonly done in analysis of single-cell RNA-seq data~\cite{haque2017practical}.
After this filtering, we retained 1,490 human cells for further analysis (\FIG{transcriptomics}B)

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested if each cell contained more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
The distribution of the amount of viral mRNA across infected cells is in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

The amount of viral mRNA was lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom).
However, viral burden remained variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
This co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS enriched IFN+ cells far beyond their initial frequency.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\cite{stetson2006type,honda2006type}.

@@

\subsection*{Validation that viral defects in single IFN+ cells often increase IFN induction}

@@

To test if the viral defects identified in single IFN+ cells directly increase IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The outcome of viral infection is extremely heterogeneous, with infected cells only sometimes activating innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Because influenza virus has a high mutation rate, many cells are infected by mutated virions.
But so far, it has been impossible to fully characterize the genetic properties of the virions infecting any given cell, since the conventional techniques of flow cytometry and single-cell RNA-seq only detect if a protein or transcript is present---not its sequence.
Here we develop a new approach that uses long-read PacBio sequencing to determine the full sequences of the virions infecting individual cells.
Most cells are infected by virions that have a mutation or some other defect.
This viral genetic variation explains a substantial amount of the cell-to-cell variability in how much viral mRNA is produced in infected cells, and helps determine whether a cell activates an innate-immune response.
However, our results also show that viral genetic defects do not completely explain cell-to-cell variation in the outcome of infection.
Overall, this study provides the first complete picture of how viral mutations affect the outcome of infection in single cells.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Infection with an acute virus such as influenza initiates a race between the virus and immune system.
As the virus spreads, some cells detect infection and begin producing interferon (IFN).
This IFN directs expression of anti-viral interferon-stimulated genes (ISGs) in the infected cell and its neighbors via autocrine and paracrine signaling, as well as helping launch a broader immune response~\cite{stetson2006type,honda2006type}.
If innate immunity is activated sufficiently rapidly, it can reduce viral replication and disease~\cite{solov1969results,treanor1987intranasally,beilharz2007protection,kugel2009intranasal,steel2010transmission}---although excessive immune responses later in infection can actually be associated with immunopathology and severe disease~\cite{la2007question, iwasaki2014innate}.

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common even if cells are infected with a nominally ``wild-type'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection*{A system to identify and enrich for rare IFN+ cells}

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large B} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

@@

\caption{
{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\cite{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \SUPPFIG{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with our prior work~\cite{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\cite{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene.

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}B) among infected cells.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\cite{stetson2006type,honda2006type}.

@@

\subsection*{Validation that viral defects in single IFN+ cells often cause increased IFN induction}

@@

To test if the viral defects identified in single IFN+ cells play a causal role in triggering IFN expression, we used reverse genetics to generate bulk stocks of viruses with some of these defects.

The viral defect most strongly associated with IFN induction was failure to express the NS gene (\FIG{genotypes}, \FIG{mutations}C).
Although it is sometimes possible to use complementing cells to generate influenza viruses lacking a specific gene~\cite{fujii2003selective,marsh2007specific}, we were unable to generate viruses that lacked NS.
The NS gene encodes two proteins (NS1 and NS2), the first of which is influenza's primary innate-immune antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
We therefore mimicked the effect of absence of NS by creating a mutant virus (which we will term ``NS1stop'') that had multiple stop codons early in the NS1 coding sequence.
